\Large Inferential Analysis of the Supreme Court Citation Network

### 2018-03-23 Tex/paper.tex
updated result table 2

\maketitle
\begin{abstract}
\noindent The significance and influence of US Supreme Court majority opinions derive in large part to opinions' roles as precedents for future opinions. A growing body of literature seeks to understand what drives the use of opinions as precedents through the study of Supreme Court case citation patterns. We raise two limitations of existing work on Supreme Court citations. First, dyadic citations are typically aggregated to the case level before they are analyzed. Second, citaitons are treated as if they arise independent of one another. We make the case that studying Supreme Court citations as a dynamic citation network overcomes both of these limitations. We present a methodology for studying citations between US Supreme Court opinions at the dyadic level, as a network. This methodology---the citation temporal exponential random graph model---enables researchers to account for the effects of case characteristics and complex forms of network dependence in citation formation. We apply this methodology to a network that includes all Supreme Court cases decided between 1937 and 2001. We find evidence for dependence processes, including reciprocity, transitivity, popularity, and activity. The dependence effects that we identify are as substantively and statistically significant as the effects of the exogenous covariates we include in the model. The summary result from this analysis is that theoretical models of Supreme Court citation formation should consider both the effects of case characteristics and the structure of past citations.\footnote{
Prepared for presentation at the ``The Enemy of My Enemy: Judicial Politics meets Network Analysis'' conference within a conference at the 2018 Annual Conference of the Southern Political Science Association. This work was supported in part by NSF grants SES- 1558661, SES-1637089, SES-1619644, and CISE-1320219.}
\end{abstract}

@@

We incorporate a statistic to model the degree to which cases cite those that are similar in terms of the ideological positions of the justices who supported the decision. We account for this effect following citet{spriggs2001explaining}, who find that cases are more likely to be overruled when the Court is ideologically distant from the median justice in the majority coalition that decided the case. \citet{clark2010locating} estimates a latent coordinate model of Supreme Court opinions based on the network of case-to-case citations. They find that the majority opinion falls at the ideal point of the median member of the majority coalition in the case. We include a covariate term in which $X_{ij}$ is the absolute difference between the Martin-Quinn scores \citep{martin2002dynamic} of the median justices in the majority coalitions for cases $i$ and $j$. We expect this variable to have a negative effect, which would correspond to cases citing those to which they are ideologically similar.%WE COULD USE THIS DATASET AS A SECOND APPLICATION

@@

\citet{benjamin2012standing} Study the propensity for cases to be overruled and cited in other negative ways. They find that cases with majority coalitions that are large and ideologically broad are less likely to be cited negatively. In our data we do not differentiate between negative and positive citations, but since the overwhelming majority of citations are positive, we hypothesize that the effects they found will be reversed in our analysis. We include one covariate in which $X_{ij}$ is the number of justices in the majority coalition for case $j$. We also include another covariate in which $X_{ij}$ is the absolute difference between the maximum and minimum ideal points of the justices in the majority coalition for case $j$. We expect both of these covariates to have positive effects.

@@

Edges                                              & -5.533                           & -5.733                              & -5.388                              & *                                    \\ \hline
Instar(2)                                          & 0.031                            & 0.027                               & 0.036                               & *                                    \\ \hline
Outstar(2)                                         & 0.022                            & 0.020                               & 0.032                               & *                                    \\ \hline
Mutual                                             & 3.316                            & 2.622                               & 3.983                               & *                                    \\ \hline
Triangle                                           & 1.490                            & 1.410                               & 1.560                               & *                                    \\ \hline
Martin Quinn Score                                 & 0.080                            & 0.022                               & 0.126                               & *                                    \\ \hline
Same Issue Area Homophily                          & 1.378                            & 1.313                               & 1.451                               & *                                    \\ \hline
Year Difference                                    & -0.077                           & -0.090                              & -0.064                              & *                                    \\ \hline
$(\text{Year Differnce})^2$                        & 0.0029                           & 0.0025                              & 0.0032                              & *                                    \\ \hline
Receiver Abs Diff MQ Score in Majority             & -0.030                           & -0.048                              & -0.014                              & *                                    \\ \hline
Receiver Number Justices in Majority               & -0.115                           & -0.135                              & -0.089                              & *                                    \\ \hline
Receiver  Sender Year                              & 0.0007                           & 0.0006                              & 0.0009                              & *                                    \\ \hline
Sender Same Issue Area 2                           & 0.162                            & 0.105                               & 0.217                               & *                                    \\ \hline
Sender Same Issue Area 3                           & -0.317                           & -0.403                              & -0.236                              & *                                    \\ \hline
Sender Same Issue Area 4                           & 0.497                            & 0.426                               & 0.570                               & *                                    \\ \hline
Sender Same Issue Area 5                           & 0.332                            & 0.174                               & 0.495                               & *                                    \\ \hline
Sender Same Issue Area 6                           & 0.527                            & 0.402                               & 0.635                               & *                                    \\ \hline
Sender Same Issue Area 7                           & 0.327                            & 0.251                               & 0.394                               & *                                    \\ \hline
Sender Same Issue Area 8                           & 0.076                            & 0.021                               & 0.129                               & *                                    \\ \hline
Sender Same Issue Area 9                           & 0.261                            & 0.209                               & 0.315                               & *                                    \\ \hline
Sender Same Issue Area 10                          & 0.369                            & 0.301                               & 0.432                               & *                                    \\ \hline
Sender Same Issue Area 11                          & -0.046                           & -0.233                              & 0.151                               &                                      \\ \hline
Sender Same Issue Area 12                          & 0.0073                           & -0.075                              & 0.089                               &                                      \\ \hline
Sender Same Issue Area 13                          & 0.381                            & 0.162                               & 0.564                               & *                                    \\ \hline
Sender Same Issue Area 14                          & 0.151                            & -0.019                              & 0.300                               &                                      \\ \hline
Receiver Same Issue Area 2                         & 0.231                            & 0.166                               & 0.300                               & *                                    \\ \hline
Receiver Same Issue Area 3                         & -0.101                           & -0.209                              & -0.006                              & *                                    \\ \hline
Receiver Same Issue Area 4                         & 0.516                            & 0.425                               & 0.612                               & *                                    \\ \hline
Receiver Same Issue Area 5                         & 0.351                            & 0.183                               & 0.534                               & *                                    \\ \hline
Receiver Same Issue Area 6                         & 0.461                            & 0.284                               & 0.628                               & *                                    \\ \hline
Receiver Same Issue Area 7                         & 0.361                            & 0.281                               & 0.437                               & *                                    \\ \hline
Receiver Same Issue Area 8                         & 0.153                            & 0.097                               & 0.221                               & *                                    \\ \hline
Receiver Same Issue Area 9                         & 0.279                            & 0.221                               & 0.346                               & *                                    \\ \hline
Receiver Same Issue Area 10                        & 0.492                            & 0.413                               & 0.570                               & *                                    \\ \hline
Receiver Same Issue Area 11                        & 0.247                            & 0.056                               & 0.411                               & *                                    \\ \hline
Receiver Same Issue Area 12                        & 0.339                            & 0.246                               & 0.430                               & *                                    \\ \hline
Receiver Same Issue Area 13                        & 1.060                            & 0.845                               & 1.239                               & *                                    \\ \hline
Receiver Same Issue Area 14                        & 0.557                            & 0.255                               & 0.793                               & *                                    \\ \hline
Instar(2) $\times$ Sender Year                     & -0.00025                         & -0.00034                            & -0.00017                            & *                                    \\ \hline
Outstar(2) $\times$ Sender Year                    & -0.00036                         & -0.00056                            & -0.00028                            & *                                    \\ \hline
Mutual $\times$ Sender Year                        & -0.010                           & -0.043                              & 0.029                               &                                      \\ \hline
Triangle $\times$ Sender Year                      & -0.0004                          & -0.0021                             & 0.0016                              &                                      \\ \hline
Martin Quinn Score $\times$ Sender Year            & -0.0025                          & -0.0035                             & -0.0013                             & *                                    \\ \hline
Same Issue Area $\times$ Sender Year               & -0.0058                          & -0.0076                             & -0.0043                             & *                                    \\ \hline
Year Difference $\times$ Sender Year               & 0.0004                           & 0.0002                              & 0.0007                              & *                                    \\ \hline
$\text{Year Difference}^2 \times$ Sender Year      & -0.000037                        & -0.000044                           & -0.000030                           & *                                    \\ \hline
MQ Score in Majority $\times$ Sender Year & 0.0013                           & 0.0009                              & 0.0017                              & *                                    \\ \hline
Justices in Majority $\times$ Sender Year   & 0.0029                           & 0.0023                              & 0.0035                              & *                                    \\ \hline

@@

\caption{Bootstrapped MPLE Results for the time period $1937-2001$. A '*' indicates that the 2.5th and 97.5th quantile of the variable does not include '0' and as a result is statistically significant.}

>>>>

\maketitle
\begin{abstract}
\noindent The significance and influence of US Supreme Court majority opinions derive in large part to opinions' roles as precedents for future opinions. A growing body of literature seeks to understand what drives the use of opinions as precedents through the study of Supreme Court case citation patterns. We raise two limitations of existing work on Supreme Court citations. First, dyadic citations are typically aggregated to the case level before they are analyzed. Second, citaitons are treated as if they arise independent of one another. We make the case that studying Supreme Court citations as a dynamic citation network overcomes both of these limitations. We present a methodology for studying citations between US Supreme Court opinions at the dyadic level, as a network. This methodology---the citation temporal exponential random graph model---enables researchers to account for the effects of case characteristics and complex forms of network dependence in citation formation. We apply this methodology to a network that includes all Supreme Court cases decided between 1937 and 2015. We find evidence for dependence processes, including reciprocity, transitivity, popularity, and activity. The dependence effects that we identify are as substantively and statistically significant as the effects of the exogenous covariates we include in the model. The summary result from this analysis is that theoretical models of Supreme Court citation formation should consider both the effects of case characteristics and the structure of past citations.\footnote{
Prepared for presentation at the ``The Enemy of My Enemy: Judicial Politics meets Network Analysis'' conference within a conference at the 2018 Annual Conference of the Southern Political Science Association. This work was supported in part by NSF grants SES- 1558661, SES-1637089, SES-1619644, and CISE-1320219.}
\end{abstract}

@@

We incorporate a statistic to model the degree to which cases cite those that are similar in terms of the ideological positions of the justices who supported the decision. We account for this effect following \citet{spriggs2001explaining}, who find that cases are more likely to be overruled when the Court is ideologically distant from the median justice in the majority coalition that decided the case. \citet{clark2010locating} estimates a latent coordinate model of Supreme Court opinions based on the network of case-to-case citations. They find that the majority opinion falls at the ideal point of the median member of the majority coalition in the case. We include a covariate term in which $X_{ij}$ is the absolute difference between the Martin-Quinn scores \citep{martin2002dynamic} of the median justices in the majority coalitions for cases $i$ and $j$. We expect this variable to have a negative effect, which would correspond to cases citing those to which they are ideologically similar.%WE COULD USE THIS DATASET AS A SECOND APPLICATION

@@

\citet{benjamin2012standing} study the propensity for cases to be overruled and cited in other negative ways. They find that cases with majority coalitions that are large and ideologically broad are less likely to be cited negatively. In our data we do not differentiate between negative and positive citations, but since the overwhelming majority of citations are positive, we hypothesize that the effects they found will be reversed in our analysis. We include one covariate in which $X_{ij}$ is the number of justices in the majority coalition for case $j$. We also include another covariate in which $X_{ij}$ is the absolute difference between the maximum and minimum ideal points of the justices in the majority coalition for case $j$. We expect both of these covariates to have positive effects.

\begin{itemize}
\item mention new variable: overruled cases
\end{itemize}

@@

Edges                                              & -6.249                           & -6.412                              & -6.110                              & *                                    \\ \hline
Instar(2)                                          & 0.035                            & 0.032                               & 0.038                               & *                                    \\ \hline
Outstar(2)                                         & 0.021                            & 0.019                               & 0.028                               & *                                    \\ \hline
Mutual                                             & 2.071                            & -4.157                               & 4.072                                &                                     \\ \hline
Triangle                                           & 1.507                            & 1.431                               & 1.559                               & *                                    \\ \hline
Martin Quinn Score                                 & 0.079                            & 0.029                               & 0.130                               & *                                    \\ \hline
Same Issue Area Homophily                          & 1.447                            & 1.389                               & 1.515                               & *                                    \\ \hline
Year Difference                                    & -0.068                           & -0.079                              & -0.058                              & *                                    \\ \hline
$(\text{Year Differnce})^2$                        & 0.0019                           & 0.0016                              & 0.0022                              & *                                    \\ \hline
Receiver Abs Diff MQ Score in Majority             & 0.047                           & 0.029                              & 0.063                              & *                                     \\ \hline
Receiver Number Justices in Majority               & -0.075                           & -0.095                              & -0.055                              & *                                    \\ \hline
Receiver  Sender Year                              & 0.0039                           & 0.0036                              & 0.0042                              & *                                    \\ \hline
Overruled Cases   & -1.043                           & -1.624                              & -0.564                              & *                                    \\ \hline
Sender Same Issue Area 2                           & 0.142                            & 0.100                               & 0.185                               & *                                    \\ \hline
Sender Same Issue Area 3                           & -0.337                           & -0.403                              & -0.270                              & *                                    \\ \hline
Sender Same Issue Area 4                           & 0.498                            & 0.438                               & 0.561                               & *                                    \\ \hline
Sender Same Issue Area 5                           & 0.182                            & 0.031                               & 0.305                               & *                                    \\ \hline
Sender Same Issue Area 6                           & 0.536                            & 0.422                               & 0.648                               & *                                    \\ \hline
Sender Same Issue Area 7                           & 0.370                            & 0.316                               & 0.426                               & *                                    \\ \hline
Sender Same Issue Area 8                           & 0.107                            & 0.059                               & 0.156                               & *                                    \\ \hline
Sender Same Issue Area 9                           & 0.278                            & 0.238                               & 0.329                               & *                                    \\ \hline
Sender Same Issue Area 10                          & 0.371                            & 0.317                               & 0.433                               & *                                    \\ \hline
Sender Same Issue Area 11                          & -0.030                           & -0.151                              & 0.164                               &                                      \\ \hline
Sender Same Issue Area 12                          & 0.064                           & -0.014                              & 0.130                               &                                      \\ \hline
Sender Same Issue Area 13                          & 0.608                            & 0.423                               & 0.792                               & *                                    \\ \hline
Sender Same Issue Area 14                          & 0.232                            & 0.075                              & 0.393                               & *                                      \\ \hline
Receiver Same Issue Area 2                         & 0.252                            & 0.199                               & 0.310                               & *                                    \\ \hline
Receiver Same Issue Area 3                         & -0.092                           & -0.194                              & 0.014                              &                                     \\ \hline
Receiver Same Issue Area 4                         & 0.481                            & 0.397                               & 0.570                               & *                                    \\ \hline
Receiver Same Issue Area 5                         & 0.374                            & 0.224                               & 0.519                               & *                                    \\ \hline
Receiver Same Issue Area 6                         & 0.521                            & 0.346                               & 0.652                               & *                                    \\ \hline
Receiver Same Issue Area 7                         & 0.412                            & 0.339                               & 0.487                               & *                                    \\ \hline
Receiver Same Issue Area 8                         & 0.174                            & 0.134                               & 0.232                               & *                                    \\ \hline
Receiver Same Issue Area 9                         & 0.311                            & 0.240                               & 0.361                               & *                                    \\ \hline
Receiver Same Issue Area 10                        & 0.506                            & 0.429                               & 0.577                               & *                                    \\ \hline
Receiver Same Issue Area 11                        & 0.299                            & 0.088                               & 0.432                               & *                                    \\ \hline
Receiver Same Issue Area 12                        & 0.350                            & 0.276                               & 0.445                               & *                                    \\ \hline
Receiver Same Issue Area 13                        & 0.972                            & 0.780                               & 1.135                               & *                                    \\ \hline
Receiver Same Issue Area 14                        & 0.517                            & 0.200                               & 0.740                               & *                                    \\ \hline
Instar(2) $\times$ Sender Year                     & -0.00026                         & -0.00031                            & -0.00021                            & *                                    \\ \hline
Outstar(2) $\times$ Sender Year                    & -0.00031                         & -0.00045                            & -0.00027                            & *                                    \\ \hline
Mutual $\times$ Sender Year                        & 0.110                           & 0.010                              & 0.524                               & *                                      \\ \hline
Triangle $\times$ Sender Year                      & -0.00022                          & -0.0016                             & 0.0013                              &                                      \\ \hline
Martin Quinn Score $\times$ Sender Year            & -0.0019                          & -0.0030                             & -0.0009                             & *                                    \\ \hline
Same Issue Area $\times$ Sender Year               & -0.0068                          & -0.0081                             & -0.0055                             & *                                    \\ \hline
Year Difference $\times$ Sender Year               & 0.00046                           & 0.00028                              & 0.00065                              & *                                    \\ \hline
$\text{Year Difference}^2 \times$ Sender Year      & -0.000024                        & -0.000028                           & -0.000021                           & *                                    \\ \hline
MQ Score in Majority $\times$ Sender Year & -0.00053                           & -0.00093                              & -0.0018                              & *                                    \\ \hline
Justices in Majority $\times$ Sender Year   & 0.0017                           & 0.0013                              & 0.0021                              & *                                    \\ \hline
Overruled Cases $\times$ Sender Year   & 0.012                           & 0.002                              & 0.023                              & *                                    \\ \hline

@@

\caption{Bootstrapped MPLE Results for the time period $1937-2015$. A '*' indicates that the 2.5th and 97.5th quantile of the variable does not include '0' and as a result is statistically significant.}
